---
title: "How to use package spatcart"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{How-to-use-this-package}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(spatcart)
```

# Package description

**spatcart** provides classification trees adapted to spatial bivariate marked point processes by using a splitting criterion associated with the Kcross interaction between marks proposed by Ripley. The package also provides heatmaps coloring partitions implied by tree leaves with respect to the interaction between marks.  

The full description of the SpatCART algorithm can be found in the article *Spatial Classification Trees*, by A. Bar-Hen, S. Gey and J.-M. Poggi, HAL 01837065 (2018).

# Main features

The main features of **spatcart** are presented with respect to the lines followed in *Spatial Classification Trees*.  


## Spatial classification trees

* **spatcart**  
* **spattree, spatprune**
* **partition.spattree**

## Simulations

* **damier, repulsion**
* **gg.partition.tree**

## Paracou data set

* **Paracou**
* **Paracoudata, Paracou.plot**




